---
title: "Public Health and Economic Consequences of Severe Weather Events"
output:
  html_document:
    keep_md: TRUE
---

## SYNOPSIS

This study examines the NOAA storm event database to determine which types of weather events are the most harmful to human populations and economically.  After normalizing the data, analyses were based on the data collected between 1996 and 2011, comprised of 48 categories of events.  'Population Cost' was determined by looking separately at fatality and injury counts per event, 'Economic Cost' was determined as a combination of property and crop damage totals. 

The events with the highest population cost were Excessive Heat events (highest fatalities) and Tornadoes (highest injuries, highest combinied population impact).  Floods/flash floods were also of note from a population cost standpoint.

The events with the highest economic cost were all water-related, with flooding being the most costly.  Also notable in terms of economic damage are hurricanes/typhoons and storm tides.

---


## DATA PROCESSING

Libraries required include `tm`, `stringdist`, `stringr`, `ggplot2`, `gridExtra`, and `scales`.
Local system must also have a pdf processor installed.  I used `xpdf` on mac.
```{r}
library(tm)         # PDF processing
library(stringdist) # Approximate string matching
library(stringr)    # Consistent regex
library(ggplot2)    # Plots
library(gridExtra)  # Layout of plots
library(scales)     # Non-Scientific Notation Plot Scales
```

### SOURCING DATA

First we download and extract the NOAA storm data.
```{r cache=TRUE}
# Check for data dir
if (!file.exists("data")) {
  dir.create("data")
}

# Check for data source file, download if needed.
if (!file.exists("data/noaa_storm_data.bz2")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile="data/noaa_storm_data.bz2", method="curl")
}

noaa <- read.csv('data/noaa_storm_data.bz2',
                 header = TRUE, stringsAsFactors = FALSE, strip.white=TRUE)
dim(noaa)
```

Also let's grab the documentation from http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf.
Linux/Mac package 'xpdf' is required to extract the 48 categories. (credit to 
https://class.coursera.org/repdata-035/forum/thread?thread_id=35 for the idea).
```{r}
if (!file.exists("data/pd01016005curr.pdf")) {
  download.file("http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf", destfile="data/pd01016005curr.pdf", method="curl")
}

# Read the PDF using `tm`'s `readPDF`.  This requires a pdf engine, I used
# xpdf on a mac.
pdf <- readPDF(control = list(c(text = "-layout")))
pdf <- pdf(elem=list(uri="data/pd01016005curr.pdf"),language="en")
```

### PROCESSING CATEGORY DATA

We extract the 48 'official' categories from the table of contents in the source document.
These are matched by their format: 7.(1 or two digit number), space, text, space, event designator in parens.  Then we strip the leading/trailing metainformation.
```{r cache = TRUE}
# Extract the 'Event Types' from the table of contents
toc <- paste(pdf$content[78:91], collapse = ' ')

# Then extract the 48 'official' category names.
# We match them based on their format: 7.(1 or two digit number), 
# space, text, space, event designator in parens.
categories <- str_extract_all(toc, "7\\.[:digit:]+\\.? ([^\\)]*) \\([CMZ]\\)")[[1]]

# Then we strip the leading/trailing metainformation
categories <- str_replace(categories, "^7\\.[:digit:]+\\.? ", "")
categories <- str_replace(categories, "[:space:]+\\([CMZ]\\)$", "")

print(categories)
```

### PROCESSING STORM EVENT DATA
We'll first separate the data into groups by the collection year.
We need to coerce a date.  For our purposes, we'll determine the event's year
by the beginning date of the event.

Then we'll assign a 'collection category' corresponding to the information from here:
http://www.ncdc.noaa.gov/stormevents/details.jsp.  The categories are

- `tornado`: Tornado only
- `ttws1`: Tornado, Thunderstorm Wind and Hail 1 (manually keyed)
- `ttws2`: Tornado, Thunderstorm Wind and Hail 2 (text files)
- `all`: All event types

```{r cache=TRUE}
noaa$event_date <- as.Date(strptime(noaa$BGN_DATE, "%m/%d/%Y %H:%M:%S"))
noaa$event_year <- strftime(noaa$event_date, "%Y")
noaa$collection_category <- apply(noaa, 1, function(row){
  y <- as.numeric(row['event_year'])
  if (y < 1955) { 'tornado_only' }
  else if (y >= 1955 & y < 1993) { 'ttws1' }
  else if (y >= 1992 & y < 1996) { 'ttws2' }
  else if (y >= 1996) { 'all' }
})

table(noaa$collection_category)
```

Now let's work on shrinking our working dataset a little bit.

We start by removing the columns we don't care about from the dataset.
```{r cache=TRUE}
dim(noaa)
object.size(noaa)

noaa <- subset(noaa, select = c('event_year', 
                                'collection_category', 
                                'PROPDMG', 'PROPDMGEXP',
                                'CROPDMG', 'CROPDMGEXP',
                                'FATALITIES', 'INJURIES',
                                'EVTYPE'))
dim(noaa)
object.size(noaa)
```

We're going to disregard the tornado-only data, since it's not likely to be very
useful in a comparison (there being nothing in those years' data to compare to).
We'll retain the post-1955 data.
```{r cache=TRUE}
noaa <- noaa[noaa$collection_category != 'tornado_only', ]
dim(noaa)
object.size(noaa)
```

We also don't care about entries with no 'population' or 'economic' cost,
so in this case, if the record doesn't have values for any of 'PROPDMG',
'CROPDMG', 'FATALITIES', 'INJURIES', we'll let it go.

We'll store this one off to a separate variable, because relative frequency
of impactful events may be interesting (even if not all of those events cause
damage).  We don't need or want to normalize the categories for occurrences that
are ONLY for nonzero economic/personal cost observations.
```{r}
noaa_lim <- noaa[rowSums(noaa[ , c('PROPDMG', 'CROPDMG', 'FATALITIES', 'INJURIES')]) > 0, ]
dim(noaa_lim)
object.size(noaa_lim)
```

There are WAY TOO MANY event types listed for the remaining data.
```{r}
length(unique(noaa_lim$EVTYPE))
```

We try to match some of them up to the 48 categories.
```{r}
# exact match, not very good.
table(noaa_lim$EVTYPE %in% categories)

# case normalized string match.  better, but still not very good.
table(tolower(noaa_lim$EVTYPE) %in% tolower(categories))
```

H/T https://class.coursera.org/repdata-035/forum/thread?thread_id=8#post-284 for the idea
to use amatch to correct for typographical errors.  Assuming a max typo/whitespace difference of 3 characters, we see the following
```{r}
table(amatch(tolower(noaa_lim$EVTYPE), tolower(categories), maxDist = 3, nomatch = -1))
```

We partially populate the table, so we're working with fewer records going forward.
```{r}
noaa_lim$category <- categories[amatch(tolower(noaa_lim$EVTYPE), tolower(categories), maxDist = 3)]
```

There are still a bunch of unmatched entries

```{r cache=TRUE}
head(sort(table(noaa_lim[is.na(noaa_lim$category), ]$EVTYPE), decreasing=TRUE), 20)
```

but we can clean some/most of these up.
thunderstorm/tstm wind is the VAST majority of these, and some of the others
are relatively easy to match.  We'll match all the ones that occur with some
frequency, and some of the ones that are easy to determine.  We avoid matching things
that combine categories, or are nondeterminant (like SNOW, it could possibly fit in several 
of the specified categories)

The process used to match these is as follows:

-  Any `EVTYPE` with 'thunderstorm' or 'tstm', and without 'marine' = `Thunderstorm Wind`
-  Any `EVTYPE` with both 'flash' and 'flood' = `Flash Flood`
-  Any remaining `EVTYPE`s with 'flood' = `Flood`
-  Any `EVTYPE` with both 'wild' and 'fire' = `Wildfire`
-  Any `EVTYPE` with both 'ext' and 'cold' = `Extreme Cold/Wind Chill`
-  Any `EVTYPE` with 'landslide' = `Debris Flow`
-  Any `EVTYPE` with 'storm surge' = `Storm Tide`
-  Any `EVTYPE` with either 'hurricane' or 'typhoon' = `Hurricane/Typhoon`
-  Any `EVTYPE` with 'hail' = `Hail`
-  Any `EVTYPE` with either 'winter' or 'light snow' = `Winter Weather`

```{r cache=TRUE}
# update thunderstorm wind to include anything non-marine with thunderstorm/tstm
# (other tstorm related issues would be coded as lightning or tornado damage)
noaa_lim$category <- ifelse( is.na(noaa_lim$category) & 
                         grepl("(tstm|thunderstorm)", tolower(noaa_lim$EVTYPE)) & 
                         !grepl("marine", tolower(noaa_lim$EVTYPE)), 
                         "Thunderstorm Wind", noaa_lim$category)

# Manually take on marine t-storm wind.
noaa_lim$category <- ifelse(is.na(noaa_lim$category) & grepl("marine tstm wind", tolower(noaa_lim$EVTYPE)), "Marine Thunderstorm Wind", noaa_lim$category)

# if it matches both 'flash' and 'flood', it's a flash flood
noaa_lim$category <- ifelse(is.na(noaa_lim$category) & grepl("flash", tolower(noaa_lim$EVTYPE)) & grepl("flood", tolower(noaa_lim$EVTYPE)), "Flash Flood", noaa_lim$category)

# if it matches 'flood' (or stream fld), it's a 'regular' flood
noaa_lim$category <- ifelse(is.na(noaa_lim$category) & grepl("(flood|stream fld)", tolower(noaa_lim$EVTYPE)), "Flood", noaa_lim$category)

# wildfire
noaa_lim$category <- ifelse(is.na(noaa_lim$category) & grepl("wild", tolower(noaa_lim$EVTYPE)) & grepl("fire", tolower(noaa_lim$EVTYPE)), "Wildfire", noaa_lim$category)

# extreme cold
noaa_lim$category <- ifelse(is.na(noaa_lim$category) & grepl("ext", tolower(noaa_lim$EVTYPE)) & grepl("cold", tolower(noaa_lim$EVTYPE)), "Extreme Cold/Wind Chill", noaa_lim$category)

# Hurricanes / Tsunamis
noaa_lim$category <- ifelse(is.na(noaa_lim$category) & grepl("(hurricane|typhoon)", tolower(noaa_lim$EVTYPE)), "Hurricane/Typhoon", noaa_lim$category)

# landslides = debris flow
noaa_lim$category <- ifelse(is.na(noaa_lim$category) & grepl("landslide", tolower(noaa_lim$EVTYPE)), "Debris Flow", noaa_lim$category)

# storm surge = storm tide
noaa_lim$category <- ifelse(is.na(noaa_lim$category) & grepl("storm surge", tolower(noaa_lim$EVTYPE)), "Storm Tide", noaa_lim$category)

# hail
noaa_lim$category <- ifelse(is.na(noaa_lim$category) & grepl("hail", tolower(noaa_lim$EVTYPE)), "Hail", noaa_lim$category)

# probably safe to say 'winter' is winter weather.  We'll allow 'light snow' here too.
noaa_lim$category <- ifelse(is.na(noaa_lim$category) & grepl("(winter|light snow)", tolower(noaa_lim$EVTYPE)), "Winter Weather", noaa_lim$category)
```

This takes care of all the uncategorized event types with more than
100 observations (and nonzero economic/population cost)
```{r}
head(sort(table(noaa_lim[is.na(noaa_lim$category), ]$EVTYPE), decreasing=TRUE), 20)
```

And leaves less than 0.4% of the observations we're interested in outside
of one of the 48 major categories.
```{r}
table(is.na(noaa_lim$category))['TRUE'] / nrow(noaa_lim) * 100
```

Let's apply the process to the non-limited dataset, so we'll have accurate totals for
occurrences of each of the types when we look at freqeuencies.

```{r cache=TRUE}
noaa$category <- categories[amatch(tolower(noaa$EVTYPE), tolower(categories), maxDist = 3)]
noaa$category <- ifelse( is.na(noaa$category) &
                         grepl("(tstm|thunderstorm)", tolower(noaa$EVTYPE)) &
                         !grepl("marine", tolower(noaa$EVTYPE)),
                         "Thunderstorm Wind", noaa$category)
noaa$category <- ifelse(is.na(noaa$category) & grepl("marine tstm wind", tolower(noaa$EVTYPE)), "Marine Thunderstorm Wind", noaa$category)
noaa$category <- ifelse(is.na(noaa$category) & grepl("flash", tolower(noaa$EVTYPE)) & grepl("flood", tolower(noaa$EVTYPE)), "Flash Flood", noaa$category)
noaa$category <- ifelse(is.na(noaa$category) & grepl("(flood|stream fld)", tolower(noaa$EVTYPE)), "Flood", noaa$category)
noaa$category <- ifelse(is.na(noaa$category) & grepl("wild", tolower(noaa$EVTYPE)) & grepl("fire", tolower(noaa$EVTYPE)), "Wildfire", noaa$category)
noaa$category <- ifelse(is.na(noaa$category) & grepl("ext", tolower(noaa$EVTYPE)) & grepl("cold", tolower(noaa$EVTYPE)), "Extreme Cold/Wind Chill", noaa$category)
noaa$category <- ifelse(is.na(noaa$category) & grepl("(hurricane|typhoon)", tolower(noaa$EVTYPE)), "Hurricane/Typhoon", noaa$category)
noaa$category <- ifelse(is.na(noaa$category) & grepl("landslide", tolower(noaa$EVTYPE)), "Debris Flow", noaa$category)
noaa$category <- ifelse(is.na(noaa$category) & grepl("storm surge", tolower(noaa$EVTYPE)), "Storm Tide", noaa$category)
noaa$category <- ifelse(is.na(noaa$category) & grepl("hail", tolower(noaa$EVTYPE)), "Hail", noaa$category)
noaa$category <- ifelse(is.na(noaa$category) & grepl("(winter|light snow)", tolower(noaa$EVTYPE)), "Winter Weather", noaa$category)
```

---

## RESULTS

Since we're looking to determine the economic and human impacts of storm events, we should begin by determining how we'll measure each of these and preparing those data.

### Human Impact

For `most harmful with respect to population health`, we'll take a look at both fatalities and injuries.  We'll aggregate the totals by category over the entire time period, and facet on collection period (to see if the lower number of categories in the `ttws` periods taints the outcome)

```{r}
# collect the overall totals for facet ordering
af <- aggregate(noaa$FATALITIES, 
                     by = list('category' = noaa$category), 
                     sum)
af <- af[order(af$x, decreasing = TRUE), ]
af$category <- factor(af$category, levels = af[order(af$x, decreasing = TRUE), ]$category)

# then collect the faceted set for plotting
aggregate_fatalities <- aggregate(noaa$FATALITIES, 
                                  by = list('category' = noaa$category,
                                            'collection_category' = noaa$collection_category), 
                                  sum)

aggregate_fatalities$collection_category <- factor(aggregate_fatalities$collection_category, levels = c('ttws1', 'ttws2', 'all'))
aggregate_fatalities$category <- factor(aggregate_fatalities$category, levels = af$category)

# Do the same for injuries
ai <- aggregate(noaa$INJURIES, 
                     by = list('category' = noaa$category), 
                     sum)
ai <- ai[order(ai$x, decreasing = TRUE), ]
ai$category <- factor(ai$category, levels = ai[order(ai$x, decreasing = TRUE), ]$category)

aggregate_injuries   <- aggregate(noaa$INJURIES,
                                  by = list('category' = noaa$category,
                                            'collection_category' = noaa$collection_category), 
                                  sum)
aggregate_injuries$collection_category <- factor(aggregate_fatalities$collection_category, levels = c('ttws1', 'ttws2', 'all'))
aggregate_injuries$category <- factor(aggregate_injuries$category, levels = ai$category)
```

```{r cache=TRUE}
top20_fatalities_faceted <- ggplot(aggregate_fatalities[aggregate_fatalities$category %in% head(af$category, 20), ], aes(x = category, y = x, fill = x)) + geom_bar(stat = 'identity') + scale_fill_gradient2(mid='lightgrey', high='darkred') + coord_flip() + facet_wrap(~ collection_category) + ggtitle('Fatalities by Category') + ylab('Fatality Count')

top20_injuries_faceted <- ggplot(aggregate_injuries[aggregate_injuries$category %in% head(ai$category, 20), ], aes(x = category, y = x, fill = x)) + geom_bar(stat = 'identity') + scale_fill_gradient2(mid='lightgrey', high='darkred') + coord_flip() + facet_wrap(~ collection_category) + ggtitle('Injuries by Category') + ylab('Injury Count')

grid.arrange(top20_fatalities_faceted, top20_injuries_faceted, ncol = 1)
```

From this we can see that the ttws1 period heavily skews our outcomes toward tornadoes (which were likely logged much more frequently than other weather events during the period).  What's more, not much interesting/useful (comparably) comes from ttws2.  So for the purposes of this examination, we'll focus on the 'all' period (1996 onwards).

Let's re-plot just the 'all observation' periods:

```{r cache=TRUE}
# let's repeat with only the 'all' period
noaa_all <- noaa[noaa$collection_category == 'all', ]

af_all <- aggregate(noaa_all$FATALITIES, 
                    by = list('category' = noaa_all$category), 
                    sum)
af_all$category <- factor(af_all$category, levels = af_all[order(af_all$x, decreasing = TRUE), ]$category)

ai_all <- aggregate(noaa_all$INJURIES, 
                    by = list('category' = noaa_all$category), 
                    sum)
ai_all$category <- as.character(ai_all$category)
ai_all$category <- factor(ai_all$category, levels = ai_all[order(ai_all$x, decreasing = TRUE), ]$category)

# And produce plots for just the 'all'
top20_fatalities <- ggplot(head(af_all[order(af_all$x, decreasing = TRUE), ], 20), aes(x = category, y = x, fill = x)) + geom_bar(stat = 'identity') + scale_fill_gradient2(mid='lightgrey', high='darkred') + coord_flip() + ggtitle('Fatalities by Category\n(post-1996 collection period)')  + ylab('Fatality Count')

top20_injuries <- ggplot(head(ai_all[order(ai_all$x, decreasing = TRUE), ], 20), aes(x = category, y = x, fill = x)) + geom_bar(stat = 'identity') + scale_fill_gradient2(mid='lightgrey', high='darkred') + coord_flip() + ggtitle('Injuries by Category\n(post-1996 collection period)') + ylab('Injury Count')

grid.arrange(top20_fatalities, top20_injuries, ncol = 1)
```

We can observe from these plots that the highest fatality count is from Excessive Heat, whereas Tornadoes produce far and away the highest number of injuries.  Since tornadoes also produce a very high level of fatalities, any recommendation to focus on particular events for population cost would do well to examine and prepare for Tornadoes and Excessive Heat, and possibly also Flooding/Flash Flooding.

---

### Economic Impact

For economic impact, we'll combine property (`PROPDMG`) and crop (`CROPDMG`) damage into a total dollar value per incident.

In the specification document (https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf, section 2.7) it is indicated that the numeric values of damage estimates are rounded to 3 significant digits, then abbreviated by a magnitude indicator

> Alphabetical characters used to signify magnitude include “K” for thousands, “M” for millions, and “B” for billions.

This magnitude indicator is in the `PROPDMGEXP` and `CROPDMGEXP` variables.  Some of the magnitude indicators don't follow the alpha guidelines, and so we don't know how to normalize them:

```{r cache=TRUE}
# Total nonzero property damage event counts
nonzero_prop_counts <- table(noaa$PROPDMG > 0)
print(nonzero_prop_counts)

# Table of event counts with nonzero property damage and invalid magnitude indicator
invalid_prop_magnitude_counts <- addmargins(table(noaa[noaa$PROPDMG > 0 & !grepl('[kKmMbB ]', noaa$PROPDMGEXP), ]$PROPDMGEXP))
print(invalid_prop_magnitude_counts)

# Percent of events with nonzero property damage that have an invalid magnitude indicator
(invalid_prop_magnitude_counts['Sum'] / nonzero_prop_counts['TRUE']) * 100

# Total nonzero crop damage event counts
nonzero_crop_counts <- table(noaa$CROPDMG > 0)
print(nonzero_crop_counts)

# Table of event counts with nonzero crop damage and invalid magnitude indicator
invalid_crop_magnitude_counts <- addmargins(table(noaa[noaa$CROPDMG > 0 & !grepl('[kKmMbB ]', noaa$CROPDMGEXP), ]$CROPDMGEXP))
print(invalid_crop_magnitude_counts)

# Percent of events with nonzero crop damage that have an invalid magnitude indicator
(invalid_crop_magnitude_counts['Sum'] / nonzero_crop_counts['TRUE']) * 100
```

Since for both variables these comprise a proportionally very small number of incidents (both less than 0.15 percent), we'll simply exclude these incidents from our value calculations.

To normalize the values, we'll decompress and combine the values for property and crop damage, and combine them into a single total dollar value for each incident.

```{r cache = TRUE}
# Populate a column with total economic cost per incident
noaa$economic_cost <- apply(noaa, 1, function(row){
  pval <- as.numeric(row['PROPDMG'])
  if (pval == 0 | !grepl('[kKmMbB ]', row['PROPDMGEXP'])){
    pval <- 0
  } else if (grepl('^[kK]$', row['PROPDMGEXP'])) {
    pval <- pval * 1000
  } else if (grepl('^[mM]$', row['PROPDMGEXP'])) {
    pval <- pval * 1000000
  } else if (grepl('^[bB]$', row['PROPDMGEXP'])) {
    pval <- pval * 1000000000
  }

  cval <- as.numeric(row['CROPDMG'])
  if (cval == 0 | !grepl('[kKmMbB ]', row['CROPDMGEXP'])){
    cval <- 0
  } else if (grepl('^[kK]$', row['CROPDMGEXP'])) {
    cval <- cval * 1000
  } else if (grepl('^[mM]$', row['CROPDMGEXP'])) {
    cval <- cval * 1000000
  } else if (grepl('^[bB]$', row['CROPDMGEXP'])) {
    cval <- cval * 1000000000
  }

  pval + cval
})
```

Now we can aggregate economic cost by event type and compare. Since we've already determined there is heavy skew in the pre-1996 data, we'll be examining only the post-1996 data here.
After aggregating, we order the output to be highest aggregate cost to lowest.
We also convert the event type to a factor and force-reorder the levels in the
same arrangement (for plotting purposes).

```{r cache=TRUE}
noaa_all <- noaa[noaa$collection_category == 'all', ]

aggregate_cost <- aggregate(noaa_all$economic_cost, 
                            by = list('category' = noaa_all$category), 
                            sum)
aggregate_cost <- aggregate_cost[order(aggregate_cost$x, decreasing = TRUE), ]
head(aggregate_cost, 25)

# Order the categories for the plot
aggregate_cost$category <- factor(aggregate_cost$category, levels = aggregate_cost[order(aggregate_cost$x, decreasing = TRUE), ]$category)

```

Let's take a look at the event types and see if there's a clear pattern
```{r}
top_25 <- ggplot(head(aggregate_cost[order(aggregate_cost$x, decreasing = TRUE), ], 20), aes(x = category, y = x, fill = x)) + geom_bar(stat = 'identity') + scale_fill_gradient2(mid='lightgrey', high='darkred') + coord_flip() + ggtitle('Total Economic Cost\n(post-1996 collection period)')  + ylab('Cost in Dollars') + scale_y_continuous(labels = comma)

print(top_25)
```

From this analysis, it's pretty clear that Flooding is the most economically harmful weather event type.  Additionally, Hurricanes and Storm Tides (possibly also related to hurricanes or tropical storms) account for a large amount of total economic cost.

```{r}
occ <- as.data.frame(table(noaa_all$category))
occ$percent <- round((occ$Freq / nrow(noaa_all)) * 100, 2)
occ[order(occ$percent, decreasing = TRUE), ]
```
